---
title: "Power Calculations"
teaching: 0
exercises: 0
questions:
- "?"
objectives:
- ""
- ""
keypoints:
- "."
- "."
- "."
- "."
source: Rmd
---

```{r load_libraries, include=FALSE}
source("../bin/chunk-options.R")
knitr_fig_path("08-")
library(tidyverse)
```

## Power Calculations 

#### Introduction
We saw that there is always a certain level of variation and uncertainty around our estimates due to sampling, and we saw how to quantify it and account for it when we're trying to learn something about our population. We also saw that in order to be able to overcome that variation and uncertainty and detect any effect in our data, we need to have a large enough sample size. Another way to say it is that we need to have enough statistical power to detect a real effect if it exists. What constitutes a large enough sample size will change from one study to the next because power depends on both the amppling variation and the effect size in a specific study. It also depends on the p-value threshold we choose, which is basically our tolerance for reaching the wrong conclusion. 
So it is often useful to calculate the power in our study in order to optimize our sample size or at least be able to say whether or not we have a large enough sample size. Of course, more data is always better in terms of detecting real effects, but it comes with a cost of money and time, as well as exposing subjects to suffering and potential health risks. Power calculation helps us optimize this trade-off. 

Below we will see how to calculate power and how the different factors mentioned above - sampling variation, effect size, and p-value cut off - affect our ability to statistically detect real effects.

To see it in action:
We have used the example of the effects of two different diets on body weight of mice. Since in this illustrative example we have access to the "real" population, we know that in fact there is a substantial (about 10%) difference between the average weights of the two populations:

```{r, message=FALSE}
library(dplyr)
pheno <- read.csv("../data/mice_pheno.csv") #Previously downloaded 

controlPopulation <- filter(pheno, Sex == "F" & Diet == "chow") %>%  
  select(Bodyweight) %>% unlist

hfPopulation <- filter(pheno, Sex == "F" & Diet == "hf") %>%  
  select(Bodyweight) %>% unlist

mu_hf <- mean(hfPopulation)
mu_control <- mean(controlPopulation)
print(mu_hf - mu_control)
print((mu_hf - mu_control)/mu_control * 100) #percent increase
```

We have also seen that, in some cases, when we take a sample and perform a 
t-test, we don't always get a p-value smaller than 0.05. For example, here is a 
case where we take a sample of 5 mice and don't achieve statistical significance 
at the 0.05 level:

```{r}
set.seed(1234)
N <- 5
hf <- sample(hfPopulation, N)
control <- sample(controlPopulation, N)
t.test(hf, control)$p.value
```

Did we make a mistake? By not rejecting the null hypothesis, are we saying the 
diet has no effect? The answer to this question is no. All we can say is that we 
did not reject the null hypothesis. But this does not necessarily imply that the 
null is true. The problem is that, in this particular instance, we don't have 
enough _power_, which is largely the same as saying that we don't have a large enough sample size for this particular analysis. 

#### Types of Error

In the example above, we couldn't reject the null hypothesis because our sample size is too small. But that decision was based on the p-value we chose to consider as our cutoff. Over the years, the convention has become to choose cutoffs of 0.05 and 0.01 in most cases, but this is ultimately a largely arbitrary - scientifically-blind - choice.  

We could decide in the example above that we accept any p-value below 0.25 to be considered significant, in which case we would be able to reject the null hypothesis with only 5 data points.

Then why wouldn't we want to do that?

Our choice of a cut-off p-value basically reflects a trade-off between two different types of errors we might make in the conclusion we draw:

A higher cutoff means that we would reject the null hypothesis more easily - with smaller sample size and for smaller effect size - regardless of whether it is true or false. That means that a higher cutoff will make us more likely to make the mistake of rejecting it when it is true, or concluding that the effect size we observe is real when it is not. This *error* is called _type I error_ by statisticians, or false positive. With p-value cutoff of 0.05, that will happen 1 out of 20 times. With cutoff of 0.25 that will happen 1 out of 4 times. 

But a smaller cutoff, say 0.01, means that we are more likely to make the opposite mistake of not rejecting the null hypothesis when it is in fact wrong. This is called a _type II error_ or a false negative. 
The R code analysis above shows an example of a false negative: we did not reject the null hypothesis (at the 0.05 level) and, because we happen to know and peeked at the true population means, we know there is in fact a difference. Had we used a p-value cutoff of 0.25, we would not have made this mistake. However, in general, are we comfortable with a type I error rate of 1 in 4? Usually we are not.  

Power calculation helps us understand the level of risk we are taking in each direction, with different combinations of sample size and effect size.

#### Power Calculation

Power is the probability of rejecting the null when the null is false, i.e., the probability of correctly detecting a real effect in our dataset. 

Power depends on the magnitude of the effect we're estimating - e.g., how large is the difference between two groups - and the sampling variation, which we usually measure with standard deviation and which depends on sample size.

In practice, we don't know the true standard deviation and we certainly don't know in advance what is the real effect size. So we want to look at several plausible combinations of effect size and sample size.

As with everything else, we can either use simulations to get those insights or make assumptions about the probability distribution and then use the appropriate formulas that statistical theory gives us. 
Here we will illustrate the concepts behind 
power by coding up simulations in R. 

Suppose our sample size is: 

```{r}
N <- 12
```

and we will reject the null hypothesis at:

```{r}
alpha <- 0.05
```

What is our power with this particular data? We will compute this probability by 
re-running the exercise many times and calculating the proportion of times the 
null hypothesis is rejected. Specifically, we will run: 

```{r}
B <- 2000
```

The simulation is as follows: we take a sample of size *N* from 
both control and treatment groups, we perform a t-test comparing these two, and 
report if the p-value is less than `alpha` or not. We write a function that does 
this:

```{r}
reject <- function(N, alpha=0.05){
   treatment <- sample(hfPopulation, N) 
   control <- sample(controlPopulation, N)
   pval <- t.test(treatment, control)$p.value
   pval < alpha
}
```

Here is an example of one simulation for a sample size of 12. The call to 
`reject` answers the question "Did we reject?" 

```{r}
reject(N=12)
```

Now we can use the `replicate` function to do this `B` times. 

```{r}
rejections <- replicate(B, reject(N))
```

Our power is just the proportion of times we correctly reject. So with  *N = 12* 
our power is only: 

```{r}
length(which(rejections==TRUE))/B
```

This explains why the t-test was not rejecting the null when we knew it was false. 
With a sample size of just 12, our power is about 23%. This is considered a very low power. Usually we want to aim for at least 80%.

Let's see how power improves with N. We will use the function `sapply`, which 
applies a function to each of the elements of a vector. We want to repeat the 
above for the following sample size:

```{r}
Ns <- seq(5, 50, 5)
```

So we use `apply` like this:
```{r}
power <- sapply(Ns,function(N){
  rejections <- replicate(B, reject(N))
  length(which(rejections==TRUE))/B
  })
```

For each of the simulations, the above code returns the proportion of 
times we reject. Not surprisingly power increases with N:

```{r power_versus_sample_size, fig.cap="Power plotted against sample size."}
plot(Ns, power, type="b")
```

Similarly, if we change the level `alpha` at which we reject, power changes. The 
smaller I want the chance of type I error to be, the less power I will have. 
Another way of saying this is that we trade off between the two types of error. 
We can see this by writing similar code, but keeping *N* fixed and considering 
several values of `alpha`:

```{r power_versus_alpha, fig.cap="Power plotted against cut-off."}
N <- 30
alphas <- c(0.1, 0.05, 0.01, 0.001, 0.0001)
power <- sapply(alphas, function(alpha){
  rejections <- replicate(B, reject(N, alpha=alpha))
  length(which(rejections==TRUE))/B
})
plot(alphas, power, xlab="alpha", type="b", log="x")
```

Note that the x-axis in this last plot is in the log scale.

There is no "right" power or "right" alpha level, but it is important that you 
understand what each means.

To see this clearly, you could create a plot with curves of power versus N. Show 
several curves in the same plot with color representing alpha level.


#### p-values Are Arbitrary under the Alternative Hypothesis

Another consequence of what we have learned about power is that p-values are 
somewhat arbitrary when the null hypothesis is not true and therefore the 
*alternative* hypothesis is true (the difference between the population means is 
not zero). When the alternative hypothesis is true,  we can make a p-value as 
small as we want simply by increasing the sample size (supposing that we have an 
infinite population to sample from). We can show this property of p-values by 
drawing larger and larger samples from our population and calculating p-values. 
This works because, in our case, we know that the alternative hypothesis is 
true, since we have access to the populations and can calculate the difference 
in their means.

First write a function that returns a p-value for a given sample size *N*:

```{r}
calculatePvalue <- function(N) {
   hf <- sample(hfPopulation,N) 
   control <- sample(controlPopulation,N)
   t.test(hf,control)$p.value
}
```

We have a limit here of 200 for the high-fat diet population, but we can see the 
effect well before we get to 200. For each sample size, we will calculate a few 
p-values. We can do this by repeating each value of *N* a few times.

```{r}
Ns <- seq(10,200,by=10)
Ns_rep <- rep(Ns, each=10)
```

Again we use `sapply` to run our simulations:

```{r}
pvalues <- sapply(Ns_rep, calculatePvalue)
```

Now we can plot the 10 p-values we generated for each sample size:

```{r pvals_decrease, fig.cap="p-values from random samples at varying sample size. The actual value of the p-values decreases as we increase sample size whenever the alternative hypothesis is true."}
plot(Ns_rep, pvalues, log="y", xlab="sample size",
     ylab="p-values")
abline(h=c(.01, .05), col="red", lwd=2)
```

Note that the y-axis is log scale and that the p-values show a decreasing trend 
all the way to 10<sup>-8</sup> as the sample size gets larger. The standard 
cutoffs of 0.01 and 0.05 are indicated with horizontal red lines.

It is important to remember that p-values are not more interesting as they 
become very very small. Once we have convinced ourselves to reject the null 
hypothesis at a threshold we find reasonable, having an even smaller p-value 
just means that we sampled more mice than was necessary.  Having a larger sample 
size does help to increase the precision of our estimate of the difference 
<i>&Delta;</i>, but the fact that the p-value becomes very very small is just a 
natural consequence of the mathematics of the test.  The p-values get smaller 
and smaller with increasing sample size because the numerator of the t-statistic
has <i>&radic;N</i> (for equal sized groups, and a similar effect occurs
when <i>M &ne; N</i>). Therefore, if <i>&Delta;</i> is non-zero, the t-statistic
will increase with <i>N</i>.

Therefore, a better statistic to report is the effect size with a confidence 
interval or some statistic which gives the reader a sense of the change in a 
meaningful scale. We can report the effect size as a percent by dividing the 
difference and the confidence interval by the control population mean:

```{r}
N <- 12
hf <- sample(hfPopulation, N)
control <- sample(controlPopulation, N)
diff <- mean(hf) - mean(control)
diff / mean(control) * 100
t.test(hf, control)$conf.int / mean(control) * 100
```

In addition, we can report a statistic called
[Cohen's d](https://en.wikipedia.org/wiki/Effect_size#Cohen.27s_d),
which is the difference between the groups divided by the pooled standard
deviation of the two groups. 

```{r}
sd_pool <- sqrt(((N-1) * var(hf) + (N-1) * var(control))/(2 * N - 2))
diff / sd_pool
```

This tells us how many standard deviations of the data the mean of the high-fat 
diet group is from the control group. Under the alternative hypothesis, unlike 
the t-statistic which is guaranteed to increase, the effect size and Cohen's d 
will become more precise.

> ## Exercises
> For these exercises we will load the babies dataset from babies.txt. We will 
> use this data to review the concepts behind the p-values and then test 
> confidence interval concepts.  
> `url<-"https://raw.githubusercontent.com/genomicsclass/dagdata/master/inst/extd\ata/babies.txt"`  
> `filename <- basename(url)`  
> `download(url, destfile=filename)`  
> `babies<-read.table("babies.txt", header=TRUE)`  
> This is a large dataset (1,236 cases), and we will pretend that it contains 
> the entire population inwhich we are interested. We will study the differences 
> in birth weight between babies born to smoking and non-smoking mothers.First, 
> let’s split this into two birth weight datasets: one of birth weights to 
> non-smoking mothers and the other of birth weights to smoking mothers.  
> `bwt.nonsmoke <- filter(babies, smoke==0) %>%`  
> `select(bwt) %>%`  
> `unlist`  
> `bwt.smoke<-filter(babies, smoke==1) %>%`  
> `select(bwt) %>%`  
> `unlist`   
> Now, we can look for the true population difference in means between smoking 
> and non-smoking birth weights.  
> `library(rafalib)`  
> `mean(bwt.nonsmoke) - mean(bwt.smoke)`  
> `popsd(bwt.nonsmoke)`  
> `popsd(bwt.smoke)`  
> The population difference of mean birth weights is about 8.9 ounces. The 
> standard deviations of the nonsmoking and smoking groups are about 17.4 and 
> 18.1 ounces, respectively.As we did with the mouse weight data, this 
> assessment interactively reviews inference concepts using simulations in R. We
> will treat the babies dataset as the full population and draw samples from it 
> to simulate individual experiments. We will then ask whether somebody who only 
> received the random samples would be able to draw correct conclusions about 
> the population. We are interested in testing whether the birth weights of 
> babies born to non-smoking mothers are significantly different from the birth 
> weights of babies born to smoking mothers.  
> 1. Set the seed at 1 and obtain two samples, each of size <i>N = 25</i>, from 
> non-smoking mothers (`dat.ns`) and smoking mothers (`dat.s`). Compute the 
> t-statistic (call it tval).  
> 2. Recall that we summarize our data using a t-statistic because we know that 
> in situations where the null hypothesis is true (what we mean when we say 
> “under the null”) andthe sample size is relatively large, this t-value will 
> have an approximate standard normal distribution. Because we know the 
> distribution of the t-value under the null, we can quantitatively determine 
> how unusual the observed t-value would be if the null hypothesis were true.The 
> standard procedure is to examine the probability a t-statistic that actually 
> does follow the null hypothesis would have larger absolute value than the 
> absolute value of the t-value we just observed – this is called a two-sided 
> test. We have computed these by taking one minus the area under the standard 
> normal curve between `-abs(tval)` and `abs(tval)`. In R, we can do this by 
> using the `pnorm` function, which computes the area under a normal curve from 
> negative infinity up to the value given as its first argument:  
> 3. Because of the symmetry of the standard normal distribution, there is a 
> simpler way to calculate the probability that a t-value under the null could 
> have a larger absolute value than `tval`. Choose the simplified calculation 
> from the following:  
> A) 1 - 2 * pnorm(abs(tval))  
> B) 1 - 2 * pnorm(-abs(tval))  
> C) 1 - pnorm(-abs(tval))  
> D) 2 * pnorm(-abs(tval))  
> 4. By reporting only p-values, many scientific publications provide an 
> incomplete story of their findings. As we have mentioned, with very large 
> sample sizes, scientifically insignificant differences between two groups can 
> lead to small p-values. Confidence intervals are more informative as they 
> include the estimate itself. Our estimate of the difference between babies of 
> smoker and non-smokers: `mean(dat.s)  -  mean(dat.ns)`. If we use the CLT, 
> what quantity would we add and subtract to this estimate to obtain a 99% 
> confidence interval?  
> 5. If instead of CLT, we use the t-distribution approximation, what do we add 
> and subtract (use 2 * N - 2 degrees of freedom)?  
> 6. Why are the values from 4 and 5 so similar?  
> A) Coincidence.  
> B) They are both related to 99% confidence intervals.  
> C) N and thus the degrees of freedom is large enough to make the normal and 
> t-distributions very similar.  
> D) They are actually quite different, differing by more than 1 ounce.  
> 7. No matter which way you compute it, the p-value `pval` is the probability 
> that the null hypothesis could have generated a t-statistic more extreme than
> what we observed: `tval`. If the p-value is very small, this means that 
> observing a value more extreme than `tval` would be very rare if the null 
> hypothesis were true, and would give strong evidence that we should reject the 
> null hypothesis. We determine how small the p-value needs to be to reject the 
> null by deciding how often we would be willing to mistakenly reject the null 
> hypothesis.  
> The standard decision rule is the following: choose some small 
> value <i>&alpha;</i> (in most disciplines the conventional choice is 
> <i>&alpha;</i> 0:05) and reject the null hypothesis if the p-value is less
> than <i>&alpha;</i>. We call the significance level of the test.  
> It turns out that if we follow this decision rule, the probability that we 
> will reject the null hypothesis by mistake is equal to <i>&alpha;</i>. (This 
> fact is not immediately obvious and requires some probability theory to show.)
> We call the event of rejecting the null hypothesis, when it is in fact true, a 
> Type I error, we call the probability of making a Type I error, the Type I 
> error rate, and we say that rejecting the null hypothesis when the p-value is 
> less than <i>&alpha;</i>, controls the Type I error rate so that it is equal 
> to <i>&alpha;</i>. We will see a number of decision rules that we use in order 
> to control the probabilities of other types of errors. Often, we will 
> guarantee that the probability of an error is less than some level, but, in 
> this case, we can guarantee that the probability of a Type I error is exactly 
> equal to <i>&alpha;</i>.  
> Which of the following sentences about a Type I error is not true?  
> A) The following is another way to describe a Type I error: you decided to 
> reject the null hypothesis on the basis of data that was actually generated by 
> the null hypothesis.  
> B) The following is the another way to describe a Type I error: due to random
> fluctuations, even though the data you observed were actually generated by the 
> null hypothesis, the p-value calculated from the observed data was small, so 
> you rejected it.  
> C) From the original data alone, you can tell whether you have made a Type I 
> error.  
> D) In scientific practice, a Type I error constitutes reporting a 
> “significant” result when there is actually no result.  
> 8. In the simulation we have set up here, we know the null hypothesis is false 
> – the true value of difference in means is actually around 8.9. Thus, we are 
> concerned with how often the decision rule outlined in the last section allows
> us to conclude that the null hypothesis is actually false. In other words, we 
> would like to quantify the Type II error rate of the test, or the probability 
> that we fail to reject the null hypothesis when the alternative hypothesis is
> true.  
> Unlike the Type I error rate, which we can characterize by assuming that the 
> null hypothesis of “no difference” is true, the Type II error rate cannot be 
> computed by assuming the alternative hypothesis alone because the alternative 
> hypothesis alone does not specify a particular value for the difference. It 
> thus does not nail down a specific distribution for thet-value under the 
> alternative.  
> For this reason, when we study the Type II error rate of a hypothesis testing 
> procedure, we need to assume a particular effect size, or hypothetical size of 
> the difference between population means, that we wish to target. We ask 
> questions such as “what is the smallest difference I could reliably 
> distinguish from 0 given my sample size N?” or, more commonly,“How big does N
> have to be in order to detect that the absolute value of the difference is 
> greater than zero?” Type II error control plays a major role in designing data 
> collection procedures before you actually see the data, so that you know the 
> test you will run has enough sensitivity or power. Power is one minus the Type 
> II error rate, or the probability that you will reject the null hypothesis 
> when the alternative hypothesis is true.  
> There are several aspects of a hypothesis test that affect its power for a 
> particular effect size. Intuitively, setting a lower <i>&alpha;</i> decreases 
> the power of the test for a given effect size because the null hypothesis will 
> be more difficult to reject. This means that for an experiment with fixed 
> parameters (i.e., with a predetermined sample size, recording mechanism, etc), 
> the power of the hypothesis test trades off with its Type I error rate, no 
> matter what effect size you target.  
> We can explore the trade off of power and Type I error concretely using the 
> babies data. Since we have the full population, we know what the true effect 
> size is (about 8.93) and we can compute the power of the test for true 
> difference between populations.  
> Set the seed at 1 and take a random sample of N = 5 measurements from each of 
> the smoking and nonsmoking datasets. What is the p-value (use the `t-test` 
> function)?  
> 9. The p-value is larger than 0.05 so using the typical cut-off, we would not 
> reject. This is a type II error. Which of the following is not a way to 
> decrease this type of error?  
> A) Increase our chance of a type I error.  
> B) Take a larger sample size  
> C) Find a population for which the null is not true.  
> D) Use a higher level.  
> 10. Set the seed at 1, then use the `replicate` function to repeat the code 
> used in exercise 9 10,000 times. What proportion of the time do we reject at 
> the 0.05 level?  
> 11. Note that, not surprisingly, the power is lower than 10%. Repeat the 
> exercise above for samples sizes of 30, 60, 90 and 120. Which of those four 
> gives you power of about 80%?  
> 12. Repeat problem 11, but now require an <i>&alpha;</i> level of 0.01. Which 
> of those four gives you power of about 80%?
>
> > ## Solution to Exercises
> {: .solution}
{: .challenge}
